---
title: "Sleep-Quality"
author: 
  - name: "Savaira Manoa"
  - name: "Adrian Pouflis"
  - name: "Eduard Vargas"
affiliation: "**Monash Student ID: 34322450**"
date: "2024-06-01"
output: 
  bookdown::html_document2:
    number_sections: false
  bookdown::pdf_document2:
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```



#### Section 1: Loading the libraries
```{r Loading the libraries, include=FALSE}
# Loading the libraries
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(scales)
library(RColorBrewer)
library(viridis)
library(htmltools)

```


#### Research Question:
What is the impact of various lifestyle choices on sleep quality among adults?

#### Introduction to the data set.

The  data set utilized for this work was obtained from [Kaggle](https://www.kaggle.com/datasets/equilibriumm/sleep-efficiency).

Kaggle is a website and collection of services for data science. It provides a platform for users to find and share datasets, as well as compete in machine learning competitions. Since the business _"strongly encourage dataset publishers to share their data in an accessible, non-proprietary format if possible"_, the sleep quality data set does mentions the license is under the Authors, for which this case is Equilibrium, but there is no more information to be found in the Equilibrium's Bio. Nonetheless, this data set has a 10.0 out of 10.0 grade on usability rated by Kaggle users, and it can be clearly identified that the data set has been used multiple times by different users for academic purposes.


Within the Kaggle's data set page, the introduction displays that the dataset contains information about a group of test subjects and their sleep patterns, including sleep duration, efficiency, and stages (REM, deep, and light sleep), as well as factors like caffeine intake, alcohol consumption, smoking status, and exercise frequency, in addition to age and gender.




#### Data Set Introduction:

```{r Loading Data}
## Loading the data set
raw_data <- read.csv("Sleep_Efficiency.csv",header = TRUE)
```


The following are the variables identified in the data set.
```{r Looking at variables}
# Identifying the variables:

print(colnames(raw_data))
```

As the intention of this collaborative project is to pin point the relevant results with regards to : _"What is the impact of various lifestyle choices on sleep quality among adults?"_, then instead of creating an overall analysis, we will focus on determined variables and their impact upon sleeping quality.


**Explanation of the Raw Data:**
```{r}
print(str(raw_data))
print(summary(raw_data))
```

 - **Analysis of the data encountered:**
 
 The data set provides comprehensive information on sleep quality and related factors for a group of 452 individuals. The dataset includes demographic information such as age, which ranges from 9 to 69 years with an average of 40.29 years, and gender. Detailed sleep metrics are recorded, including bedtime, wakeup time, total sleep duration (ranging from 5 to 10 hours with an average of 7.47 hours), and sleep efficiency (ranging from 50% to 99% with an average of 78.89%). Additionally, the dataset measures the percentage of sleep spent in different stages: REM sleep (15% to 30%), deep sleep (18% to 75%), and light sleep (7% to 63%). The number of awakenings during the night is also recorded, with values ranging from 0 to 4 times, and an average of 1.64 times.

Beyond sleep-specific data, the dataset captures lifestyle factors that may influence sleep quality. These factors include caffeine consumption (ranging from 0 to 200 mg), alcohol consumption (0 to 5 units), smoking status, and exercise frequency (0 to 5 days per week). The dataset contains some missing values in the columns for awakenings, caffeine consumption, alcohol consumption, and exercise frequency. Overall, this dataset is well-suited for investigating the relationships between sleep quality and various demographic and lifestyle factors, offering valuable insights into how different behaviors and characteristics impact sleep patterns and efficiency.


**Important metrics to understand:**

As mentioned by the Author in the Intro to the data:


- _"The "Sleep efficiency" feature is a measure of the proportion of time spent in bed that is actually spent asleep."_

- _"The "REM sleep percentage", "Deep sleep percentage", and "Light sleep percentage" features indicate the amount of time each subject spent in each stage of sleep."_

- _"The "Awakenings" feature records the number of times each subject wakes up during the night. "_




### Gender and excercise variables:


In order to understand the data a sleeping efficiency, which was defined by the author as ***"A measure of the proportion of time spent in bed that is actually spent asleep."***, we decided to use a simple linear regression model in order to find the relationships between the habits of an adult person, and the consequences on the mentioned sleeping efficiency.


Here, the intercept, or dependent variable will be "Sleep Efficiency":


```{r Cleaning data}
# Clean data by removing rows with missing values
cleaned_data <- na.omit(raw_data)

# Fit a linear model to predict Sleep.efficiency
Linear_model <- lm(Sleep.efficiency ~ Age + Caffeine.consumption + Alcohol.consumption + Exercise.frequency + Gender, data = cleaned_data)

# Get a summary of the model
summary(Linear_model)
```


**Observations:**

The linear model analysis reveals several significant relationships between the predictors and sleep efficiency. The model explains about 25.84% of the variability in sleep efficiency (R-squared = 0.2584). Key predictors such as age, alcohol consumption, and exercise frequency have significant effects on sleep efficiency, while caffeine consumption is not a significant predictor. The intercept indicates the baseline sleep efficiency when all predictors are at zero, and the overall model is statistically significant (p-value < 2.2e-16).

**Specific Observations:**

 - Gender:
Coefficient for Gender (Male): -0.0219210
Interpretation: On average, males have a sleep efficiency that is approximately 2.19% lower than females, although this effect is only marginally significant (p = 0.094462).
Significance: The marginal significance suggests that while there is a difference in sleep efficiency between genders, it is not as strongly supported by the data compared to other predictors.

- Exercise Frequency
Coefficient for Exercise Frequency: 0.0261673

**Interpretation:** Each additional day of exercise per week is associated with a 2.62% increase in sleep efficiency, holding other factors constant.

**Significance:** This relationship is highly significant (p = 1.72e-09), indicating a strong positive impact of exercise on sleep efficiency.

**Summary:**
The linear model highlights that while gender has a marginal impact on sleep efficiency, exercise frequency has a robust positive effect. These insights can inform strategies to improve sleep quality, particularly emphasizing the benefits of regular exercise.




After understanding the relevant variables that affect the "Sleeping Efficiency" it is important to visualize how the current information is, and then proceed to work on more relationships with the help of the model elaborated:

```{r Visualizing the data Gender-Excercise}
gender_exercise <- cleaned_data %>%
    select(Gender, Sleep.efficiency, Exercise.frequency)


summary_by_gender <- gender_exercise %>%
    group_by(Gender) %>%
    summarise(
        Min_Sleep_Efficiency = min(Sleep.efficiency, na.rm = TRUE),
        Max_Sleep_Efficiency = max(Sleep.efficiency, na.rm = TRUE),
        Mean_Sleep_Efficiency = mean(Sleep.efficiency, na.rm = TRUE)
    )

```

```{r Printint the tables}
kable(summary_by_gender, caption = "Summary of Sleep Efficiency by Gender")
```

Now let's have a look at the distribution of exercise frequency 

```{r Histogram visualizing the distribution, fig.align='right', fig.align='center'}
# Create histograms for each exercise frequency, faceted by gender
ggplot(gender_exercise, aes(x = Sleep.efficiency, fill = Gender)) +
  geom_histogram(binwidth = 0.05, alpha = 0.6, position = "identity") +
  facet_wrap(~ Exercise.frequency) +
  labs(title = "Distribution of Sleep Efficiency by Exercise Frequency and Gender",
       x = "Sleep Efficiency",
       y = "Count") +
  theme_minimal()

```



Now, if we were to plot the information for sleeping quality vs exercise levels by gender, it would look like this:

```{r Plotting the relationship, fig.align='center'}
Gender_Exercise_SE_Plot <- 
    ggplot(gender_exercise, aes(x = Exercise.frequency, y = Sleep.efficiency, color = Gender)) +
    geom_point()+
    labs( title = "Relationship between excercise and Sleep Efficiency by Gender",
          x = "Exercise Frequency",
          y = "Sleep Efficiency",
          color = "Gender") +
    theme_minimal()

print(Gender_Exercise_SE_Plot)
```


### Age and alcohol consumption analysis:
-Adrian Pouflis

**Research Question**

As identified in Eduard's earlier analysis, using linear regression, age and alcohol consumption both have significant impacts on sleep efficiency. Alcohol consumption had a significantly larger magnitude of impact than age, its direction was negative.

I have two primary research questions relating to these variables.

How do these variables impact quality of sleep independently?

Does one variable have a mediating effect on the other? (For example, maybe the impact of alcohol increases with age.)

**Analysis - Question 1**

With regards to the first research question, a relationship between the variables and sleep efficiency has already been quantified. For each point of alcohol consumption, sleep efficiency decreases by 3.36%. Age seems to actually have an extremely small impact, a less than 1% (~0.1664%) increase in sleep efficiency with age. Both results were found to be statistically significant. 

These effects are fairly moderate in scope, as was noted already, the linear regression explains only about 25.84% of the variance in sleep efficiency. Repeating the process on other test variables (Awakenings and sleep duration) below, similarly low R-squared values can be seen.

```{r}
Linear_model_duration <- lm(Sleep.duration ~ Age + Caffeine.consumption + Alcohol.consumption + Exercise.frequency + Gender, data = cleaned_data)

duration_r_squared = summary(Linear_model_duration)$r.squared


Linear_model_awakenings <- lm(Awakenings ~ Age + Caffeine.consumption + Alcohol.consumption + Exercise.frequency + Gender, data = cleaned_data)

awakening_r_squared = summary(Linear_model_awakenings)$r.squared

cat("\nThe r-squared value for the linear model on sleep duration is",duration_r_squared)

cat("\nThe r-squared value for the linear model on number of awakenings is",awakening_r_squared)

```
Alcohol consumption ranged between doses of 0 and 5. Let's examine sleep efficiency for each value of alcohol consumption.

```{r Visualizing the data Gender-Excercise}
alcohol_data <- cleaned_data %>%
    select(Sleep.efficiency, Alcohol.consumption)


summary_by_alcohol <- alcohol_data %>%
    group_by(Alcohol.consumption) %>%
    summarise(
        Mean_Sleep_Efficiency = mean(Sleep.efficiency, na.rm = TRUE)
    )

summary_by_alcohol
```

Overall, the relationship between alcohol consumption and sleep efficiency seems uncertain, lower amounts of alcohol do see higher efficiency, but the value spikes at a value of 3 and the regression indicates the effect is somewhat weak. 

Repeating the process with age:

```{r Visualizing the data Gender-Excercise}
Age_bands <- cut(cleaned_data$Age, c(seq(9, 70, by = 5), Inf), include.lowest = TRUE)

aggregate(Sleep.efficiency ~ Age_bands, cleaned_data, mean)
```
Sleep efficiency would appear to peak around the 29-59 range, falling in later years and being lowest in 
adolescent years.

**Analysis - Question 2**

Now I've explored the impact of these variables independently, what remains is to see how one variable can mediate the other.

(Mediation analysis can go here)
#### Citations:

Equilibrium. (2023). *Sleep efficiency* [Data set]. Kaggle. [https://www.kaggle.com/datasets/equilibriumm/sleep-efficiency](https://www.kaggle.com/datasets/equilibriumm/sleep-efficiency)
