---
title: "Sleep-Quality"
author: 
  - name: "Savaira Manoa"
  - name: "Adrian Pouflis"
  - name: "Eduard Vargas"
affiliation: "**Monash Student ID: 34322450**"
date: "2024-06-01"
output: 
  bookdown::html_document2:
    number_sections: false
  bookdown::pdf_document2:
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```



#### Section 1: Loading the libraries
```{r Loading the libraries, include=FALSE}
# Loading the libraries
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(scales)
library(RColorBrewer)
library(viridis)
library(htmltools)
library(interactions)

```


#### Overall Research Question:
What is the impact of various lifestyle choices on sleep quality among adults?

#### Introduction to the data set.

The data set utilized for this work was obtained from Kaggle, a website and collection of services for data science, including dataset hosting. The data set license is under the Authors, listed as 'Equilibrium', but the data's source was not available. Regardless, the data set is highly rated and widely used for academic purposes.


#### Data Set Introduction:

```{r Loading Data}
## Loading the data set
raw_data <- read.csv("Sleep_Efficiency.csv",header = TRUE)
```


The following are the variables identified in the data set.
```{r Looking at variables}
# Identifying the variables:

print(colnames(raw_data))
```

As the intention of this collaborative project is to answer the question of : _"What is the impact of various lifestyle choices on sleep quality among adults?"_, we will determine and analyse how a select few key variables impact sleep quality.


**Explanation of the Raw Data:**
```{r}
print(str(raw_data))
print(summary(raw_data))

```

 - **Analysis of the data encountered:**
 

The data set provides comprehensive information on sleep quality and related factors for 452 individuals. The data set includes demographic information such as age (9-69 years), and gender. Detailed sleep metrics recorded include bedtime, waketime, total sleep duration, number of awakenings during the night and sleep efficiency. Additionally, the dataset measures the percentage of sleep spent in different stages: REM, deep, and light sleep. The "Sleep efficiency" feature is a measure of the proportion of time spent asleep in bed.  

Beyond sleep-specific data, the dataset captures lifestyle factors that influence sleep quality. These factors include caffeine/alcohol consumption, smoking status, and exercise frequency. Overall, this dataset is well-suited for investigating the relationships between sleep quality and various demographic and lifestyle factors, offering insights into how behaviors and characteristics impact sleep patterns and efficiency.

The table above summarises important information about the variables.


### Gender and excercise variables:


In order to understand the data a sleeping efficiency, which was defined by the author as ***"A measure of the proportion of time spent in bed that is actually spent asleep."***, we decided to use a simple linear regression model in order to find the relationships between the habits of an adult person, and the consequences on the mentioned sleeping efficiency.


Here, the intercept, or dependent variable will be "Sleep Efficiency":


```{r Cleaning data}
# Clean data by removing rows with missing values
cleaned_data <- na.omit(raw_data)

# Fit a linear model to predict Sleep.efficiency
Linear_model <- lm(Sleep.efficiency ~ Age + Caffeine.consumption + Alcohol.consumption + Exercise.frequency + Gender, data = cleaned_data)

# Get a summary of the model
summary(Linear_model)
```


**Observations:**

The linear model analysis reveals several significant relationships between the predictors and sleep efficiency. The model explains about 25.84% of the variability in sleep efficiency (R-squared = 0.2584). Key predictors such as age, alcohol consumption, and exercise frequency have significant effects on sleep efficiency, while caffeine consumption is not a significant predictor. The intercept indicates the baseline sleep efficiency when all predictors are at zero, and the overall model is statistically significant (p-value < 2.2e-16).

**Specific Observations:**

 - Gender:
Coefficient for Gender (Male): -0.0219210
Interpretation: On average, males have a sleep efficiency that is approximately 2.19% lower than females, although this effect is only marginally significant (p = 0.094462).
Significance: The marginal significance suggests that while there is a difference in sleep efficiency between genders, it is not as strongly supported by the data compared to other predictors.

- Exercise Frequency
Coefficient for Exercise Frequency: 0.0261673

**Interpretation:** Each additional day of exercise per week is associated with a 2.62% increase in sleep efficiency, holding other factors constant.

**Significance:** This relationship is highly significant (p = 1.72e-09), indicating a strong positive impact of exercise on sleep efficiency.

**Summary:**
The linear model highlights that while gender has a marginal impact on sleep efficiency, exercise frequency has a robust positive effect. These insights can inform strategies to improve sleep quality, particularly emphasizing the benefits of regular exercise.




After understanding the relevant variables that affect the "Sleeping Efficiency" it is important to visualize how the current information is, and then proceed to work on more relationships with the help of the model elaborated:

```{r Visualizing the data Gender-Excercise 4}
gender_exercise <- cleaned_data %>%
    select(Gender, Sleep.efficiency, Exercise.frequency)


summary_by_gender <- gender_exercise %>%
    group_by(Gender) %>%
    summarise(
        Min_Sleep_Efficiency = min(Sleep.efficiency, na.rm = TRUE),
        Max_Sleep_Efficiency = max(Sleep.efficiency, na.rm = TRUE),
        Mean_Sleep_Efficiency = mean(Sleep.efficiency, na.rm = TRUE)
    )

```

```{r Printint the tables}
kable(summary_by_gender, caption = "Summary of Sleep Efficiency by Gender")
```

Now let's have a look at the distribution of exercise frequency 

```{r Histogram visualizing the distribution, fig.align='right', fig.align='center'}
# Create histograms for each exercise frequency, faceted by gender
ggplot(gender_exercise, aes(x = Sleep.efficiency, fill = Gender)) +
  geom_histogram(binwidth = 0.05, alpha = 0.6, position = "identity") +
  facet_wrap(~ Exercise.frequency) +
  labs(title = "Distribution of Sleep Efficiency by Exercise Frequency and Gender",
       x = "Sleep Efficiency",
       y = "Count") +
  theme_minimal()

```



Now, if we were to plot the information for sleeping quality vs exercise levels by gender, it would look like this:

```{r Plotting the relationship, fig.align='center'}
Gender_Exercise_SE_Plot <- 
    ggplot(gender_exercise, aes(x = Exercise.frequency, y = Sleep.efficiency, color = Gender)) +
    geom_point()+
    labs( title = "Relationship between excercise and Sleep Efficiency by Gender",
          x = "Exercise Frequency",
          y = "Sleep Efficiency",
          color = "Gender") +
    theme_minimal()

print(Gender_Exercise_SE_Plot)
```

### Age-generation and exercise analysis:
This section will focus on answering two questions involving sleep efficiency, and using the two variables age grouped by generation and exercise frequency.

**Research Question**
1. Some sleep studies have shown that sleep efficiency differs between age groups. What are these differences within the data-set chosen for this study and are there any specific 'generations' that show generally better levels of sleep efficiency compared to others?
2. Given the sleep efficiency data by generation discussed in question 1, and looking at specific life-style choices such as ensuring one exercises regularly, are levels of sleep efficiency still low amongst certain age groups and why?

**Analysis - Question 1**

Sleep efficiency studies for specific age groups, have documented sleep efficiency for the aged (Desjardins et al.,2019) or sleep efficiency studies by lifespan (Kocevska et al., 2021).

**Sleep efficiency by age generation**
This subsection will first look at sleep efficiency using age grouped by generation. Using the generational age groupings (Library of Congress, n.d.) ages of subjects were catgorised.

```{r Visualizing the data generation-Exercise}
age_exercise <- cleaned_data %>%
    mutate(age_group = cut (Age, breaks = c(9,12,27,43,59,70), include.lowest = T, labels = c("Gen_Alpha","Gen_Z","Millenial","Gen_X","Boomers")))

summary_by_generation <- age_exercise %>%
    group_by(age_group) %>%
    summarise(
        Min_Sleep_Efficiency = min(Sleep.efficiency, na.rm = TRUE),
        Max_Sleep_Efficiency = max(Sleep.efficiency, na.rm = TRUE),
        Mean_Sleep_Efficiency = mean(Sleep.efficiency, na.rm = TRUE)
    )

age_ex_percentile<- age_exercise %>% 
                        group_by(age_group) %>%
                        summarise(percentile_10=quantile(Sleep.efficiency, probs = 0.10),
                                percentile_25=quantile(Sleep.efficiency, probs = 0.25),
                                percentile_50=quantile(Sleep.efficiency, probs = 0.50),
                                percentile_75=quantile(Sleep.efficiency, probs = 0.75),
                                percentile_90=quantile(Sleep.efficiency, probs = 0.90)) 
age_ex_percentile1<- age_ex_percentile %>% pivot_longer('percentile_10':'percentile_90',names_to="percentile", values_to="S_efficiency")

summary_by_gen_perc <- age_ex_percentile1 %>%
    group_by(percentile) %>%
    summarise(
        Min_Sleep_Efficiency = min(S_efficiency, na.rm = TRUE),
        Max_Sleep_Efficiency = max(S_efficiency, na.rm = TRUE),
        Mean_Sleep_Efficiency = mean(S_efficiency, na.rm = TRUE)
    )

colnames(summary_by_generation)<- c("Freq_or_Percentile","Min_Sleep_Efficiency","Max_Sleep_Efficiency", "Mean_Sleep_Efficiency")
colnames(summary_by_gen_perc)<- c("Freq_or_Percentile","Min_Sleep_Efficiency","Max_Sleep_Efficiency", "Mean_Sleep_Efficiency")

```

```{r summary of sleepefficiency by age_generation ables}

table1<- rbind(summary_by_generation, summary_by_gen_perc) %>% 
                        kbl(caption = "Summary of Sleep Efficiency by Age Generation") %>%
                        kable_classic(bootstrap_options = c("striped","hover"))
table1

```

```{r figure of sleep efficiency to age_generation}

fig1_age<- ggplot(age_ex_percentile1, aes(age_group, y=S_efficiency))
fig1_age + geom_line(aes(linetype = percentile, group = percentile)) +
            labs( title. = "Percentile of sleep efficiency by age generation",
                          x = "Generation",
                          y = "Sleep Efficiency (%)") +
                                scale_y_continuous(labels = scales::percent) +
                            theme_classic()

```

**Analysis - Question 2**

##### Sleep efficiency by exercise frequency
This subsection analyses sleep efficiency by exercise frequency, taking into account the generation of each individual in the dataset.

```{r Visualizing the data generation-Excercise1}
age_exercise_1 <- age_exercise %>%
    mutate(freq_exercise_group = cut (Exercise.frequency, breaks = c(0,1,3,5), include.lowest = T, labels = c("Low","Medium","High")))

exercise_fr_percentile<- age_exercise_1 %>% 
                        group_by(freq_exercise_group) %>%
                        summarise(percentile_10=quantile(Sleep.efficiency, probs = 0.10),
                                percentile_25=quantile(Sleep.efficiency, probs = 0.25),
                                percentile_50=quantile(Sleep.efficiency, probs = 0.50),
                                percentile_75=quantile(Sleep.efficiency, probs = 0.75),
                                percentile_90=quantile(Sleep.efficiency, probs = 0.90)) 
exercise_fr_percentile1<- exercise_fr_percentile %>% pivot_longer('percentile_10':'percentile_90',names_to="percentile", values_to="S_efficiency")

summary_by_Exfr_perc <- exercise_fr_percentile1 %>%
    group_by(percentile) %>%
    summarise(
        Min_Sleep_Efficiency = min(S_efficiency, na.rm = TRUE),
        Max_Sleep_Efficiency = max(S_efficiency, na.rm = TRUE),
        Mean_Sleep_Efficiency = mean(S_efficiency, na.rm = TRUE)
    )

summary_by_Exfrequency <- age_exercise_1 %>%
    group_by(freq_exercise_group) %>%
    summarise(
        Min_Sleep_Efficiency = min(Sleep.efficiency, na.rm = TRUE),
        Max_Sleep_Efficiency = max(Sleep.efficiency, na.rm = TRUE),
        Mean_Sleep_Efficiency = mean(Sleep.efficiency, na.rm = TRUE)
    )

#standardise column names to use rbind and merge results of percentile and exercise frequency
colnames(summary_by_Exfrequency)<- c("Freq_or_Percentile","Min_Sleep_Efficiency","Max_Sleep_Efficiency", "Mean_Sleep_Efficiency")
colnames(summary_by_Exfr_perc)<- c("Freq_or_Percentile","Min_Sleep_Efficiency","Max_Sleep_Efficiency", "Mean_Sleep_Efficiency")

```

```{r summary of sleepefficiency by exercise frequency tables}

table2<- rbind(summary_by_Exfrequency, summary_by_Exfr_perc) %>% 
                        kbl(caption = "Summary of Sleep Efficiency by Exercise Frequency") %>%
                        kable_classic(bootstrap_options = c("striped","hover"))
table2

```



```{r figure of sleep efficiency to exercise frequency by generation}


age_exercise_1 %>%
    ggplot(aes(x=freq_exercise_group, y=Sleep.efficiency,group = freq_exercise_group)) +
                geom_abline() +
                geom_jitter(width = 0.1, height=0.1) +
                    facet_wrap(~age_group, scales = "fixed") +
                        labs( title. = "Sleep efficiency by exercise frequency & generation",
                          x = "Exercise frequency",
                          y = "Sleep efficiency")
```


### Age and alcohol consumption analysis:

- Adrian Pouflis

**Research Question**

As identified earlier using regression, age and alcohol consumption have significant impacts on sleep efficiency. Alcohol consumption had a significantly larger magnitude of impact than age, its direction was negative.

I have two research questions:

1:How do these variables impact quality of sleep independently?

2:How do these variables interact? Does one variable impact the magnitude of (moderate) the effect of the other? 

**Analysis - Question 1**

These effects are small, as was noted already, the linear regression explains only about 25.84% of the variance in sleep efficiency. Below the regression is performed on different sleep measures, to similar results. 

```{r}
Linear_model_duration <- lm(Sleep.duration ~ Age + Caffeine.consumption + Alcohol.consumption + Exercise.frequency + Gender, data = cleaned_data)

duration_r_squared = summary(Linear_model_duration)$r.squared


Linear_model_awakenings <- lm(Awakenings ~ Age + Caffeine.consumption + Alcohol.consumption + Exercise.frequency + Gender, data = cleaned_data)

awakening_r_squared = summary(Linear_model_awakenings)$r.squared

cat("\nThe r-squared value for the linear model on sleep duration is",duration_r_squared)

cat("\nThe r-squared value for the linear model on number of awakenings is",awakening_r_squared)

```
I examined directly the relationship between age/alcohol consumption and sleep efficiency.
```{r Sleep efficiency against alcohol consumption}
alcohol_data <- cleaned_data %>%
    select(Sleep.efficiency, Alcohol.consumption)


summary_by_alcohol <- alcohol_data %>%
    group_by(Alcohol.consumption) %>%
    summarise(
        Mean_Sleep_Efficiency = mean(Sleep.efficiency, na.rm = TRUE)
    )

kable(summary_by_alcohol)
```

The relationship between alcohol and sleep efficiency is uncertain, lower doses of alcohol see higher efficiency, but the value spikes at dose 3.

I both tabled age bands and charted age against efficiency:

```{r Age against sleep efficiency}
Age_bands <- cut(cleaned_data$Age, c(seq(9, 70, by = 5), Inf), include.lowest = TRUE)

chart_data_frame <- aggregate(Sleep.efficiency ~ Age, cleaned_data, mean)

age_table <- aggregate(Sleep.efficiency ~ Age_bands, cleaned_data, mean)

kable(age_table)

ggplot(data=chart_data_frame,aes(x=Age, y=Sleep.efficiency)) +
                geom_line() + 
                scale_x_continuous(name = "Age", breaks=seq(9,69,5)) +
                        labs( title. = "Sleep efficiency by Age",
                          y = "Sleep efficiency")


```

Sleep efficiency appears to peak from 29-59.

**Analysis - Question 2**

Next is to determine if and how these variables interact.

I first chart these variables against each other to see if they interact directly.
```{r Age against sleep efficiency}

Alcohol_list <- list(
  '-1.14690721659585'="0",
  '-0.146907216494845'="1",
  '0.853092783505155'="2",
  '3'="3",
  '4'="4",
  '5'="5")


#Try to plot sleep efficiency for each age at each alcohol dose, facet by alcohol dose
Age_Alcohol_Consumption_Plot <- 
    ggplot(cleaned_data, aes(x = Age, y = Sleep.efficiency,color=as.factor(Alcohol.consumption))) +
    geom_line()+
    facet_wrap(as.factor(Alcohol.consumption)) +
    scale_color_manual(labels = c(0,1,2,3,4,5),values = c("blue", "red","black","yellow","purple","brown")) +
    labs( title = "Relationship between age and sleep efficiency by alcohol consumption",
          x = "Age",
          y = "Sleep Efficiency",
          color = "Alcohol Dose") +
          theme(
          strip.background = element_blank(),
          strip.text.x = element_blank()
          )

Age_Alcohol_Consumption_Plot
```
These graphs are erratic and show no clear pattern.


Next, I will use moderation analysis, a variant on the linear regression already performed.
```{r Moderation analysis between alcohol consumption and age}
# Perform regression, this time using a different formula to determine moderation effect
Moderation_model <- lm(Sleep.efficiency ~ Age + Caffeine.consumption + Alcohol.consumption + Exercise.frequency + Gender + Age * Alcohol.consumption, data = cleaned_data)
summary(Moderation_model)
#Display the moderation model as a plot
interactions::interact_plot(Moderation_model, pred = Age, modx = Alcohol.consumption)
```
The moderation effect is visualised. The different lines represent different levels of alcohol consumption, while their slope indicates the relationship between age and sleep efficiency. These lines are close to parallel, meaning the interaction effect is small.

The last variable, 'Age:Alcohol.consumption' indicates the relationship between age and sleep efficiency decreases as alcohol consumption increases.

Given the earlier findings for age, I performed the analysis again in the 29-59 range.

```{r Moderation analysis on a narrowed age band}
#Only select values with age between 20-59
Age_selected_data <- cleaned_data[cleaned_data$Age > 59 | cleaned_data$Age < 29 ,] 

# Perform regression, this time using a different formula to determine moderation effect
Moderation_model2 <- lm(Sleep.efficiency ~ Age + Caffeine.consumption + Alcohol.consumption + Exercise.frequency + Gender + Age * Alcohol.consumption, data = Age_selected_data)

summary(Moderation_model2)
#Display the moderation model as a plot
interactions::interact_plot(Moderation_model2, pred = Age, modx = Alcohol.consumption)


```

The moderation effect is stronger here, but small. 

#### Conclusion:
Using regression, Adrian found a weaker than expected link between age, alcohol consumption and several measures of sleep quality. 

When comparing variables against sleep efficiency, Age seemed to have a more significant positive effect between 29-59. Alcohol seemed to have a more significant effect than regression indicated, though its influence peaked at a middling dose. He also identified a very weak moderation effect between age and alcohol consumption.
#### Citations:

Desjardins, S., Lapierre, S., Hudon, C., & Desgagné, A. (2019). Factors involved in sleep efficiency: a population-based study of community-dwelling elderly persons. Sleep (New York, N.Y.), 42(5), 1-. https://doi.org/10.1093/sleep/zsz038

Equilibrium. (2023). *Sleep efficiency* [Data set]. Kaggle. [https://www.kaggle.com/datasets/equilibriumm/sleep-efficiency](https://www.kaggle.com/datasets/equilibriumm/sleep-efficiency)

Kocevska, D., Lysen, T. S., Dotinga, A., Koopman-Verhoeff, M. E., Luijk, M. P. C. M., Antypa, N., Biermasz, N. R., Blokstra, A., Brug, J., Burk, W. J., Comijs, H. C., Corpeleijn, E., Dashti, H. S., de Bruin, E. J., de Graaf, R., Derks, I. P. M., Dewald-Kaufmann, J. F., Elders, P. J. M., Gemke, R. J. B. J., … Tiemeier, H. (2021). Sleep characteristics across the lifespan in 1.1 million people from the Netherlands, United Kingdom and United States: a systematic review and meta-analysis. Nature Human Behaviour, 5(1), 113–122. https://doi.org/10.1038/s41562-020-00965-x

Library of Congress. (n.d.). *Doing Consumer Research:A Resource Guide/Generations* (https://www.guides.loc.gov/consumer-research/market-segments/generations)
