---
title: "Sleep-Quality"
author: 
  - name: "Savaira Manoa"
  - name: "Adrian Pouflis"
  - name: "Eduard Vargas"
affiliation: "**Monash Student ID: 34322450**"
date: "2024-06-01"
output: 
  bookdown::html_document2:
    number_sections: false
  bookdown::pdf_document2:
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```



#### Section 1: Loading the libraries
```{r Loading the libraries, include=FALSE}
# Loading the libraries
library(report)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(ggplot2)
library(scales)
library(RColorBrewer)
library(viridis)
library(htmltools)
library(interactions)
library(gridExtra)

```


#### Overall Research Question:
What is the impact of various lifestyle choices on sleep quality among adults?

#### Introduction to the data set.

The data set utilized for this work was obtained from Kaggle, a website and collection of services for data science, including dataset hosting. The data set license is under the Authors, listed as 'Equilibrium', but the data's source was not available. Regardless, the data set is highly rated and widely used for academic purposes.


#### Data Set Introduction:

```{r Loading Data}
## Loading the data set
raw_data <- read.csv("Sleep_Efficiency.csv",header = TRUE)
```


The following are the variables identified in the data set.
```{r Looking at variables}
# Identifying the variables:

print(colnames(raw_data))
```

As the intention of this collaborative project is to answer the question of : _"What is the impact of various lifestyle choices on sleep quality among adults?"_, we will determine and analyse how a select few key variables impact sleep quality.


**Explanation of the Raw Data:**
```{r}
print(str(raw_data))
print(summary(raw_data))

```

 - **Analysis of the data encountered:**
 

The data set provides comprehensive information on sleep quality and related factors for 452 individuals. The data set includes demographic information such as age (9-69 years), and gender. Detailed sleep metrics recorded include bedtime, waketime, total sleep duration, number of awakenings during the night and sleep efficiency. Additionally, the dataset measures the percentage of sleep spent in different stages: REM, deep, and light sleep. The "Sleep efficiency" feature is a measure of the proportion of time spent asleep in bed.  

Beyond sleep-specific data, the dataset captures lifestyle factors that influence sleep quality. These factors include caffeine/alcohol consumption, smoking status, and exercise frequency. Overall, this dataset is well-suited for investigating the relationships between sleep quality and various demographic and lifestyle factors, offering insights into how behaviors and characteristics impact sleep patterns and efficiency.

The table above summarises important information about the variables.


### Gender Sleep Efficiency, Sleep Duration and  excercise Frequency variables:  -- By Eduard Vargas

**Research Questions:**

1) What is the level of exercise levels presented by gender and what is the impact of it in the sleeping quality?

2) Is there a bias or preference for each gender when talking about the exercise levels (times per week exercise is done)? if so, what is it for each gender?


In order to understand the data a sleeping efficiency, which was defined by the author as ***"A measure of the proportion of time spent in bed that is actually spent asleep."***, I decided to use a simple linear regression model in order to find the relationships between the habits of an adult person, and the consequences on the mentioned sleeping efficiency.


Here, the intercept, or dependent variable will be "Sleep Efficiency":


```{r Cleaning data}
# Clean data by removing rows with missing values
cleaned_data <- na.omit(raw_data)

# Fit a linear model to predict Sleep.efficiency
Linear_model <- lm(Sleep.efficiency ~ Age + Caffeine.consumption + Alcohol.consumption + Exercise.frequency + Gender, data = cleaned_data)

# Get a summary of the model
summary(Linear_model)
```


**Observations:**

The linear model analysis reveals several significant relationships between the predictors and sleep efficiency. The model explains about 25.84% of the variability in sleep efficiency (R-squared = 0.2584). Key predictors such as age, alcohol consumption, and exercise frequency have significant effects on sleep efficiency, while caffeine consumption is not a significant predictor. The intercept indicates the baseline sleep efficiency when all predictors are at zero, and the overall model is statistically significant (p-value < 2.2e-16).

**Specific Observations:**

 - Gender:
Coefficient for Gender (Male): -0.0219210
Interpretation: On average, males have a sleep efficiency that is approximately 2.19% lower than females, although this effect is only marginally significant (p = 0.094462).
Significance: The marginal significance suggests that while there is a difference in sleep efficiency between genders, it is not as strongly supported by the data compared to other predictors.

- Exercise Frequency
Coefficient for Exercise Frequency: 0.0261673

**Interpretation:** Each additional day of exercise per week is associated with a 2.62% increase in sleep efficiency, holding other factors constant.

**Significance:** This relationship is highly significant (p = 1.72e-09), indicating a strong positive impact of exercise on sleep efficiency.

**Summary:**
The linear model highlights that while gender has a marginal impact on sleep efficiency, exercise frequency has a robust positive effect. These insights can inform strategies to improve sleep quality, particularly emphasizing the benefits of regular exercise.




After understanding the relevant variables that affect the "Sleeping Efficiency" it is important to visualize how the current information is, and then proceed to work on more relationships with the help of the model elaborated:

First of all, it is important to understand what is the average exercise level and reported sleeping quality per gender:


```{r Visualizing the data Gender-Excercise 4}
# Function to truncate to 2 decimal places:
trunc2 <- function(x) {
  floor(x * 100) / 100
}



gender_exercise <- cleaned_data %>%
    select(Gender, Sleep.efficiency, Exercise.frequency, Sleep.duration)


summary_by_gender <- gender_exercise %>%
    group_by(Gender) %>%
    summarise(
        Min_Sleep_Efficiency = trunc2(min(Sleep.efficiency, na.rm = TRUE)),
        Max_Sleep_Efficiency = trunc2(max(Sleep.efficiency, na.rm = TRUE)),
        Average_Sleep_Efficiency = trunc2(mean(Sleep.efficiency, na.rm = TRUE)),
        Average_exercise_levels = trunc2(mean(Exercise.frequency, na.rm  = TRUE))
    )

```

```{r Printint the tables}
kable(summary_by_gender, caption = "Summary of Sleep Efficiency by Gender")
```


**Observations from the table:**

After organizing the data, it is evident that despite males reporting higher sleep efficiency, females actually exhibit higher average sleep quality with less average exercise per week. This discrepancy could be associated with gender-related perceptions or ego, which do not reflect the actual evidence. The data suggests that females maintain better sleep efficiency with fewer exercise sessions compared to their male counterparts, highlighting an interesting dynamic in sleep quality assessments between genders.

In the following section, I will plot and visualize the relationships between exercise frequency, duration and sleep efficiency. To identify possible trends in sleep efficiency levels, in order to examine the distribution, I will make use of  histogram and scatter plots. These visualizations will help in understanding how exercise frequency influences sleep efficiency and whether consistent patterns emerge across different levels of physical activity.


#### Sleep Efficiency and Exercise Frequency per Gender:

```{r Histogram visualizing the distribution, fig.align='right', fig.align='center'}
# Create histograms for each exercise frequency, faceted by gender
ggplot(gender_exercise, aes(x = Sleep.efficiency, fill = Gender)) +
  geom_histogram(binwidth = 0.05, alpha = 0.6, position = "identity") +
  facet_wrap(~ Exercise.frequency) +
  labs(title = "Distribution of Sleep Efficiency by Exercise Frequency and Gender",
       x = "Sleep Efficiency",
       y = "Count") +
  theme_minimal()

```


```{r Plotting the relationship, fig.align='center'}
Gender_Exercise_SE_Plot <- 
    ggplot(gender_exercise, aes(x = Exercise.frequency, y = Sleep.efficiency, color = Gender)) +
    geom_point()+
    labs( title = "Relationship between excercise and Sleep Efficiency by Gender",
          x = "Exercise Frequency",
          y = "Sleep Efficiency",
          color = "Gender") +
    theme_minimal()

print(Gender_Exercise_SE_Plot)
```



Both the dot plot and the histogram display the distribution of sleep efficiency by exercise frequency and gender, revealing several important observations:

1) The general shape of the distribution indicates that sleep efficiency tends to skew towards higher values, with many individuals exhibiting sleep efficiencies closer to 1.0. This suggests that a significant portion of the population enjoys high sleep efficiency. Additionally, a positive relationship between exercise frequency and sleep efficiency is apparent, particularly noticeable at exercise frequencies of 3 and 5 days per week, implying that regular exercise may contribute to better sleep efficiency.

2) Breaking down the data by exercise frequency, we observe varied patterns for different levels of physical activity. For individuals who do not exercise at all (0 days), both males and females show a wide range of sleep efficiencies, with some clustering at lower efficiencies (around 0.6) and others at higher efficiencies (0.9 to 1.0). As exercise frequency increases to 1 and 2 days per week, the concentration of individuals with sleep efficiencies around 0.8 to 1.0 becomes more pronounced, although fewer individuals exercise only 2 days per week. For those exercising 3 days per week, there is a notable peak in high sleep efficiency, especially among males. The trend of high sleep efficiency continues for those exercising 4 and 5 days per week, despite the lower number of individuals in these categories.

3) When comparing genders, both males and females exhibit a trend of higher sleep efficiency with increased exercise frequency. Females tend to have a more evenly spread distribution across different exercise levels but consistently show high sleep efficiency. Males, on the other hand, display a stronger concentration at higher sleep efficiencies, particularly at higher levels of exercise frequency. Overall, the data suggests that regular exercise, particularly at moderate to high levels (3-5 days a week), is associated with better sleep efficiency for both genders, reinforcing the benefits of physical activity on sleep quality.



#### Sleep Duration, Efficiency and the relationship with Exercise Frequency per Gender:

```{r extracting data for male and female, include=FALSE}
female_data <- raw_data %>%
    filter(Gender == "Female")

female_quality_vs_duration <- ggplot(data = female_data, aes(x= Sleep.duration, y = Sleep.efficiency)) +
    geom_point(color = "red") +
    labs( title = "Sleep Duration vs Efficiency Females",
          x = "Sleep Duration",
          y = "Sleep Efficiency")
         
    theme_minimal()
    
#--------------------------------------

male_data <- raw_data %>%
    filter(Gender == "Male")
    
male_quality_vs_duration <- ggplot(data = male_data, aes(x= Sleep.duration, y = Sleep.efficiency)) +
    geom_point(color = "#84d3d3") +
    labs( title = "Sleep Duration vs Efficiency Males",
          x = "Sleep Duration",
          y = "Sleep Efficiency")
         
    theme_minimal()


```


```{r printing the plots, fig.align='center'}
grid.arrange(female_quality_vs_duration, male_quality_vs_duration, ncol = 2)
```

```{r Histogram visualizing the distribution Exercise vs Sleep Duration, fig.align='right', fig.align='center'}
# Create histograms for each exercise duration, faceted by gender
ggplot(gender_exercise, aes(x = Sleep.duration, fill = Gender)) +
  geom_histogram(binwidth = 1, alpha = 0.6, position = "identity") +
  facet_wrap(~ Exercise.frequency) +
  labs(title = "Distribution of Sleep Duration by Exercise Frequency and Gender",
       x = "Sleep Duration",
       y = "Count") +
  theme_minimal()
```

**Observations on the plots:**

These two visualizations provide a detailed analysis of the relationship between sleep duration, sleep efficiency, exercise frequency, and gender:

1) On the scatter plots at the top, for females, sleep efficiency generally remains high, above 0.8, across various sleep durations, with a noticeable concentration around 7 to 8 hours. They also indicate that a significant number of females maintain high sleep efficiency regardless of their exact sleep duration within this range. Similarly, for males, sleep efficiency is also predominantly high, though there is slightly more variability compared to females. Males exhibit a similar pattern of maintaining high sleep efficiency across different sleep durations, particularly between 6 and 9 hours.

2) The histograms at the bottom show the distribution of sleep duration across different exercise frequencies, comparing males and females. For individuals who do not exercise at all (0 days), both genders exhibit a concentration of sleep durations around 7 to 8 hours, with females showing a slightly higher count. This pattern is consistent for low exercise frequency (1 day), where the distribution is quite balanced between genders. As exercise frequency increases to moderate levels (2 days), there are fewer data points, but the trend of sleep durations around 7 to 8 hours continues, particularly for females. For moderate to high exercise frequencies (3 days), males show a higher count of sleep durations around 7 to 8 hours compared to females. For high (4 days) and very high exercise frequencies (5 days), the data is sparse, but the clustering around 7 to 8 hours persists, with a slight balance between males and females.

3) Overall, the scatter plots suggest that both males and females maintain high sleep efficiency (above 0.8) across various sleep durations, particularly between 6 and 9 hours. The histograms indicate that regardless of exercise frequency, both genders tend to have a similar distribution of sleep durations, predominantly around 7 to 8 hours. These observations highlight that regular exercise does not significantly alter the preferred sleep duration but suggests a tendency towards maintaining high sleep efficiency within these durations. The consistency across genders and exercise frequencies underscores the importance of sleep duration in achieving optimal sleep efficiency, suggesting that individuals may achieve better sleep quality by focusing on maintaining a consistent sleep duration.



#### General conclusions: 

The analysis of sleep efficiency in relation to exercise frequency and gender reveals that a significant portion of the population enjoys high sleep efficiency, particularly as exercise frequency increases. The distribution indicates that many individuals exhibit sleep efficiencies closer to 1.0, with a positive relationship between exercise frequency and sleep efficiency becoming evident, especially at frequencies of 3 and 5 days per week. This trend implies that regular exercise contributes to better sleep efficiency, reinforcing the benefits of physical activity on sleep quality. Notably, individuals who exercise more frequently tend to maintain higher sleep efficiency, highlighting the importance of an active lifestyle for achieving optimal sleep.

Further examination of the data reveals distinct patterns based on gender and exercise frequency. Both males and females show higher sleep efficiency with increased exercise, although the distribution patterns vary slightly. Females tend to have a more evenly spread distribution across different exercise levels, while males display a stronger concentration of higher sleep efficiencies, particularly at higher exercise frequencies. Regardless of exercise frequency, both genders exhibit a consistent sleep duration, predominantly around 7 to 8 hours, suggesting that exercise does not significantly alter the preferred sleep duration but enhances sleep efficiency within this range. These observations suggest that individuals may achieve better sleep quality by focusing on regular exercise and maintaining a consistent sleep duration.

This proves the significant correlation shown by the linear model at the beginning of this work.


#### Responding to the research questions:

1) **What is the level of exercise presented by gender and what is the impact of it on sleep quality?**

The level of exercise presented by gender shows that both males and females engage in varying frequencies of exercise per week, ranging from no exercise to exercising 5 days a week. The impact of exercise on sleep quality is evident in both genders. For females, high sleep efficiency (above 0.8) is maintained across different exercise frequencies, with a concentration of sleep durations around 7 to 8 hours. Similarly, males also exhibit high sleep efficiency with increased exercise frequency, particularly between 3 to 5 days per week, and maintain a sleep duration of 7 to 8 hours. This indicates that regular exercise contributes positively to sleep efficiency for both genders. However, the impact is slightly more pronounced in males, who show a stronger concentration of high sleep efficiencies at higher exercise levels.

2) **Is there a bias or preference for each gender when talking about the exercise levels (times per week exercise is done)? If so, what is it for each gender?**

Yes, there appears to be a bias or preference in exercise levels between genders. Females tend to have a more evenly spread distribution across different exercise levels but generally show a higher average sleep efficiency with fewer exercise sessions per week. Males, on the other hand, tend to exercise more frequently, with a noticeable concentration at higher exercise levels (3 to 5 days per week), and exhibit a higher sleep efficiency, particularly at these higher exercise frequencies. This suggests that while females achieve high sleep efficiency with less frequent exercise, males prefer and perhaps benefit more from higher levels of physical activity to achieve optimal sleep efficiency.

### Age-generation and exercise analysis by *Savaira Manoa*
This section will focus on answering two questions involving sleep efficiency using the two variables, age grouped by generation and exercise frequency.

**Research Questions**

1. Some sleep studies have shown that sleep efficiency differs between age groups. What are these differences within the dataset chosen for this study and are there any specific 'generations' that show generally better levels of sleep efficiency compared to others?


2. Given the sleep efficiency data by generation discussed in question 1, and looking at specific life-style choices such as ensuring one exercises regularly, are levels of sleep efficiency still low amongst certain age groups and why?

**Analysis - Question 1: Sleep efficiency by age generation **

Sleep efficiency studies for specific age groups, have documented sleep efficiency for the aged (Desjardins et al.,2019) or sleep efficiency studies by lifespan (Kocevska et al., 2021). 

This subsection will first look at sleep efficiency using age grouped by generation. Using the generational age groupings (Library of Congress, n.d.) ages of subjects were catgorised.  Generation Alpha shows results for 9-11 years, Generation Z for 12-26 years, Millenial for 27-42 years, Generation X for 43-58 years and the Baby Boomers for 59-69 years.

```{r Visualizing the data generation-Exercise}
#to arrange age data by generation and summarise

age_exercise <- cleaned_data %>%
    mutate(age_group = cut (Age, breaks = c(9,12,27,43,59,70), include.lowest = T, labels = c("Gen_Alpha","Gen_Z","Millenial","Gen_X","Boomers")))

summary_by_generation <- age_exercise %>%
    group_by(age_group) %>%
    summarise(
        Min_Sleep_Efficiency = min(Sleep.efficiency, na.rm = TRUE),
        Max_Sleep_Efficiency = max(Sleep.efficiency, na.rm = TRUE),
        Mean_Sleep_Efficiency = mean(Sleep.efficiency, na.rm = TRUE)
    )

age_ex_percentile<- age_exercise %>% 
                        group_by(age_group) %>%
                        summarise(percentile_10=quantile(Sleep.efficiency, probs = 0.10),
                                percentile_25=quantile(Sleep.efficiency, probs = 0.25),
                                percentile_50=quantile(Sleep.efficiency, probs = 0.50),
                                percentile_75=quantile(Sleep.efficiency, probs = 0.75),
                                percentile_90=quantile(Sleep.efficiency, probs = 0.90)) 
age_ex_percentile1<- age_ex_percentile %>% pivot_longer('percentile_10':'percentile_90',names_to="percentile", values_to="S_efficiency")

summary_by_gen_perc <- age_ex_percentile1 %>%
    group_by(percentile) %>%
    summarise(
        Min_Sleep_Efficiency = min(S_efficiency, na.rm = TRUE),
        Max_Sleep_Efficiency = max(S_efficiency, na.rm = TRUE),
        Mean_Sleep_Efficiency = mean(S_efficiency, na.rm = TRUE)
    )

colnames(summary_by_generation)<- c("~","Min_Sleep_Efficiency","Max_Sleep_Efficiency", "Mean_Sleep_Efficiency")
colnames(summary_by_gen_perc)<- c("~","Min_Sleep_Efficiency","Max_Sleep_Efficiency", "Mean_Sleep_Efficiency")

```

```{r }
#to show sleep efficiency averages in a table
table1<- rbind(summary_by_generation, summary_by_gen_perc) %>% 
                        kbl(align = NULL, caption = "Summary of sleep efficiency by age generation") %>%
                        kable_classic(bootstrap_options = c("striped","hover")) %>%
		pack_rows("Generation", 1, 4)  %>%
		pack_rows("Percentile", 6, 4)
table1

```

The table above shows sleep efficiency averages for each of the six generations mentioned.  The data indicates that Generation Alpha has the least sleep efficiency with an average of 0.54 which is within the lowest 10% of the data set population, while Generation X shows the best sleep efficiency at 0.81 which is close to the top 25% sleep efficiency results.  

```{r figure of sleep efficiency to age_generation}

fig1_age<- ggplot(age_ex_percentile1, aes(age_group, y=S_efficiency))
fig1_age + geom_line(aes(linetype = percentile, group = percentile, color = percentile)) +
            labs( title. = "Percentile of sleep efficiency by age generation",
                          x = "Generation",
                          y = "Sleep Efficiency (%)") +
                                scale_y_continuous(labels = scales::percent) +
                            theme_classic()

```

We explore this further in the Figure above.  The Figure shows the percentiles of each generation for sleep efficiency.  When read in conjunction with the table above, Generation Alpha has sleep efficiency in the lower 50% of the population.  On average, the Baby Boomers have 0.75 sleep efficency which represents approximately 50% of their population within the dataset, Generation X has 0.81 sleep efficiency which represents approximately less than 50% of their population, the Millenials has 0.80 sleep efficiency which also represents less than 50% of their population, and Generation Z has 0.74 sleep efficiency representing less than 50% of their population.  Therefore, on average Generation X have the highest sleep efficiency, but this represents less than half of their population. 


**Analysis - Question 2: Sleep efficiency by exercise frequency**

This subsection analyses sleep efficiency by exercise frequency, taking into account the generation of each individual in the dataset.

```{r Visualizing the data generation-Excercise1}
age_exercise_1 <- age_exercise %>%
    mutate(freq_exercise_group = cut (Exercise.frequency, breaks = c(0,1,3,5), include.lowest = T, labels = c("Low","Medium","High")))

exercise_fr_percentile<- age_exercise_1 %>% 
                        group_by(freq_exercise_group) %>%
                        summarise(percentile_10=quantile(Sleep.efficiency, probs = 0.10),
                                percentile_25=quantile(Sleep.efficiency, probs = 0.25),
                                percentile_50=quantile(Sleep.efficiency, probs = 0.50),
                                percentile_75=quantile(Sleep.efficiency, probs = 0.75),
                                percentile_90=quantile(Sleep.efficiency, probs = 0.90)) 
exercise_fr_percentile1<- exercise_fr_percentile %>% pivot_longer('percentile_10':'percentile_90',names_to="percentile", values_to="S_efficiency")

summary_by_Exfr_perc <- exercise_fr_percentile1 %>%
    group_by(percentile) %>%
    summarise(
        Min_Sleep_Efficiency = min(S_efficiency, na.rm = TRUE),
        Max_Sleep_Efficiency = max(S_efficiency, na.rm = TRUE),
        Mean_Sleep_Efficiency = mean(S_efficiency, na.rm = TRUE)
    )

summary_by_Exfrequency <- age_exercise_1 %>%
    group_by(freq_exercise_group) %>%
    summarise(
        Min_Sleep_Efficiency = min(Sleep.efficiency, na.rm = TRUE),
        Max_Sleep_Efficiency = max(Sleep.efficiency, na.rm = TRUE),
        Mean_Sleep_Efficiency = mean(Sleep.efficiency, na.rm = TRUE)
    )

#standardise column names to use rbind and merge results of percentile and exercise frequency
colnames(summary_by_Exfrequency)<- c("Freq_or_Percentile","Min_Sleep_Efficiency","Max_Sleep_Efficiency", "Mean_Sleep_Efficiency")
colnames(summary_by_Exfr_perc)<- c("Freq_or_Percentile","Min_Sleep_Efficiency","Max_Sleep_Efficiency", "Mean_Sleep_Efficiency")

```

```{r }

table2<- rbind(summary_by_Exfrequency, summary_by_Exfr_perc) %>% 
                        kbl(align=NULL, caption = "Summary of sleep efficiency by exercise frequency") %>%
                        kable_classic(bootstrap_options = c("striped","hover")) %>%
		pack_rows("Frequency", 1, 4)  %>%
		pack_rows("Percentile", 4, 4)
table2

```

The table above shows sleep efficiency by the three chosen frequencies of exercise:  low represents nil or 1 frequency, medium represents 2 or 3 and high greater than 3.  Sleep efficiency on average is higher with high exercise.  Exploring this further by generation, we look at the table above together with the Figure below.

```{r figure of sleep efficiency to exercise frequency by generation}


age_exercise_1 %>%
    ggplot(aes(x=freq_exercise_group, y=Sleep.efficiency,group = freq_exercise_group)) +
                geom_abline() +
                geom_jitter(width = 0.1, height=0.1) +
                    facet_wrap(~age_group, scales = "fixed") +
                        labs( title. = "Sleep efficiency by exercise frequency & generation",
                          x = "Exercise frequency",
                          y = "Sleep efficiency")
```

The Figure above shows by generation, which exercise frequencies are more or less popular.  Generation Alpha is the standout with low exercise which explains their low sleep efficiency, but may also be due to a lower sample number.  Generation X shows a majority of medium exercise frequency which explains the higher sleep efficiency.  The results for Millennials curiously show a clustering of over 0.8 sleep efficiency even with low exercise, which is above the average of 0.75.  This may be due to other lifestyle factors and a more resilient age.

**Conclusion**
In the Age Generation by Exercise analysis section by Savaira, we found that Generation Alpha has the least sleep efficiency, while Generation X on average have the highest sleep efficiency.  This was also inline with the results for exercise frequency,  where Generation Alpha is the standout with low exercise and low sleep efficiency and  Generation X showing a majority of medium exercise corresponding to higher sleep efficiency.  Curiously, Millennials show a clustering of above average 0.8 sleep efficiency even with low exercise, which may be due to other lifestyle factors and a more resilient age.

### Age and alcohol consumption analysis:

- Adrian Pouflis

**Research Question**

As identified earlier using regression, age and alcohol consumption have significant impacts on sleep efficiency. Alcohol consumption had a significantly larger magnitude of impact than age, its direction was negative.

I have two research questions:

1:How do these variables impact quality of sleep independently?

2:How do these variables interact? Does one variable impact the magnitude of (moderate) the effect of the other? 

**Analysis - Question 1**

These effects are small, as was noted already, the linear regression explains only about 25.84% of the variance in sleep efficiency. Below the regression is performed on different sleep measures, to similar results. 

```{r}
Linear_model_duration <- lm(Sleep.duration ~ Age + Caffeine.consumption + Alcohol.consumption + Exercise.frequency + Gender, data = cleaned_data)

duration_r_squared = summary(Linear_model_duration)$r.squared


Linear_model_awakenings <- lm(Awakenings ~ Age + Caffeine.consumption + Alcohol.consumption + Exercise.frequency + Gender, data = cleaned_data)

awakening_r_squared = summary(Linear_model_awakenings)$r.squared

cat("\nThe r-squared value for the linear model on sleep duration is",duration_r_squared)

cat("\nThe r-squared value for the linear model on number of awakenings is",awakening_r_squared)

```
I examined directly the relationship between age/alcohol consumption and sleep efficiency.
```{r Sleep efficiency against alcohol consumption}
alcohol_data <- cleaned_data %>%
    select(Sleep.efficiency, Alcohol.consumption)


summary_by_alcohol <- alcohol_data %>%
    group_by(Alcohol.consumption) %>%
    summarise(
        Mean_Sleep_Efficiency = mean(Sleep.efficiency, na.rm = TRUE)
    )

kable(summary_by_alcohol)
```

The relationship between alcohol and sleep efficiency is uncertain, lower doses of alcohol see higher efficiency, but the value spikes at dose 3.

I both tabled age bands and charted age against efficiency:

```{r Age against sleep efficiency Adrian}
Age_bands <- cut(cleaned_data$Age, c(seq(9, 70, by = 5), Inf), include.lowest = TRUE)

chart_data_frame <- aggregate(Sleep.efficiency ~ Age, cleaned_data, mean)

age_table <- aggregate(Sleep.efficiency ~ Age_bands, cleaned_data, mean)

kable(age_table)

ggplot(data=chart_data_frame,aes(x=Age, y=Sleep.efficiency)) +
                geom_line() + 
                scale_x_continuous(name = "Age", breaks=seq(9,69,5)) +
                        labs( title. = "Sleep efficiency by Age",
                          y = "Sleep efficiency")


```

Sleep efficiency appears to peak from 29-59.

**Analysis - Question 2**

Next is to determine if and how these variables interact.

I first chart these variables against each other to see if they interact directly.
```{r Age against sleep efficiency}

Alcohol_list <- list(
  '-1.14690721659585'="0",
  '-0.146907216494845'="1",
  '0.853092783505155'="2",
  '3'="3",
  '4'="4",
  '5'="5")


#Try to plot sleep efficiency for each age at each alcohol dose, facet by alcohol dose
Age_Alcohol_Consumption_Plot <- 
    ggplot(cleaned_data, aes(x = Age, y = Sleep.efficiency,color=as.factor(Alcohol.consumption))) +
    geom_line()+
    facet_wrap(cleaned_data$Alcohol.consumption) +
    scale_color_manual(labels = c(0,1,2,3,4,5),values = c("blue", "red","black","darkgreen","purple","brown")) +
    labs( title = "Relationship between age and sleep efficiency by alcohol consumption",
          x = "Age",
          y = "Sleep Efficiency",
          color = "Alcohol Dose") +
          theme(
          strip.background = element_blank(),
          strip.text.x = element_blank()
          )

Age_Alcohol_Consumption_Plot
```
These graphs are erratic and show no clear pattern.


Next, I will use moderation analysis, a variant on the linear regression already performed.
```{r Moderation analysis between alcohol consumption and age}
# Perform regression, this time using a different formula to determine moderation effect
Moderation_model <- lm(Sleep.efficiency ~ Age + Caffeine.consumption + Alcohol.consumption + Exercise.frequency + Gender + Age * Alcohol.consumption, data = cleaned_data)
summary(Moderation_model)
#Display the moderation model as a plot
interactions::interact_plot(Moderation_model, pred = Age, modx = Alcohol.consumption)
```
The moderation effect is visualised. The different lines represent different levels of alcohol consumption, while their slope indicates the relationship between age and sleep efficiency. These lines are close to parallel, meaning the interaction effect is small.

The last variable, 'Age:Alcohol.consumption' indicates the relationship between age and sleep efficiency decreases as alcohol consumption increases.

Given the earlier findings for age, I performed the analysis again in the 29-59 range.

```{r Moderation analysis on a narrowed age band}
#Only select values with age between 20-59
Age_selected_data <- cleaned_data[cleaned_data$Age > 59 | cleaned_data$Age < 29 ,] 

# Perform regression, this time using a different formula to determine moderation effect
Moderation_model2 <- lm(Sleep.efficiency ~ Age + Caffeine.consumption + Alcohol.consumption + Exercise.frequency + Gender + Age * Alcohol.consumption, data = Age_selected_data)

summary(Moderation_model2)
#Display the moderation model as a plot
interactions::interact_plot(Moderation_model2, pred = Age, modx = Alcohol.consumption)


```

The moderation effect is stronger here, but small. 


#### Conclusion:

Using regression, Adrian found a weaker than expected link between age, alcohol consumption and several measures of sleep quality. 

When comparing variables against sleep efficiency, Age seemed to have a more significant positive effect between 29-59. Alcohol seemed to have a more significant effect than regression indicated, though its influence peaked at a middling dose. He also identified a very weak moderation effect between age and alcohol consumption.


#### General Conclusions:

For both sexes, exercise positively impacts sleep efficiency. Males show a stronger concentration of high sleep efficiencies at higher exercise levels, while females achieve high sleep efficiency with fewer exercise sessions per week. Additionally, the relationship between alcohol consumption and sleep efficiency is more nuanced. Lower doses of alcohol are associated with higher sleep efficiency, but this effect peaks and diminishes with higher doses. Sleep duration also plays a crucial role; both males and females maintain high sleep efficiency with sleep durations primarily ranging between 7 to 8 hours. Regular exercise combined with moderate alcohol consumption and consistent sleep duration significantly enhances overall sleep quality across genders.
When examining sleep efficiency by generation, Generation X shows the highest sleep efficiency, correlating with medium exercise frequency, while Generation Alpha has the lowest sleep efficiency, likely due to low exercise frequency. Regular exercise significantly enhances sleep efficiency across all generations, although Generation Alpha shows lower efficiency, possibly due to fewer data points.
Regular exercise and age independently influence sleep efficiency, with higher exercise frequency and middle age (29-59) positively affecting sleep efficiency. The interaction between these variables suggests that while both positively impact sleep efficiency, males benefit more from higher exercise frequencies. This combined effect highlights the importance of maintaining a balanced lifestyle to achieve optimal sleep quality.



#### Citations:

Desjardins, S., Lapierre, S., Hudon, C., & Desgagné, A. (2019). Factors involved in sleep efficiency: a population-based study of community-dwelling elderly persons. Sleep (New York, N.Y.), 42(5), 1-. https://doi.org/10.1093/sleep/zsz038

Equilibrium. (2023). *Sleep efficiency* [Data set]. Kaggle. [https://www.kaggle.com/datasets/equilibriumm/sleep-efficiency](https://www.kaggle.com/datasets/equilibriumm/sleep-efficiency)

Kocevska, D., Lysen, T. S., Dotinga, A., Koopman-Verhoeff, M. E., Luijk, M. P. C. M., Antypa, N., Biermasz, N. R., Blokstra, A., Brug, J., Burk, W. J., Comijs, H. C., Corpeleijn, E., Dashti, H. S., de Bruin, E. J., de Graaf, R., Derks, I. P. M., Dewald-Kaufmann, J. F., Elders, P. J. M., Gemke, R. J. B. J., … Tiemeier, H. (2021). Sleep characteristics across the lifespan in 1.1 million people from the Netherlands, United Kingdom and United States: a systematic review and meta-analysis. Nature Human Behaviour, 5(1), 113–122. https://doi.org/10.1038/s41562-020-00965-x

Library of Congress. (n.d.). *Doing Consumer Research:A Resource Guide/Generations* (https://www.guides.loc.gov/consumer-research/market-segments/generations)

Package citations:

```{r Citations}
cite_packages()
```
